Multi-Domain Adaptation in Neural Machine Translation with Dynamic Sampling Strategies

### 2021-05-14 2021/emnlp/auto_cl_mdmt.tex
local changes

We now compare the best MDAC (initialized by $\lambda_0$) to fine-tuning. According to Table \ref{tab:da}, MDAC is outperformed by fine-tuning method in large domains including \domain{med} and \domain{law}. However, MDAC improves fine-tuning by approximately 0.5 BLEU in \domain{bank} and 1.0 BLEU in \domain{rel}. This indicates that for small domains, out-of-domain data can improve the generalization of the NMT model and MDAC is capable to exploit this advantage by balancing the in-domain and the out-of-domain training data instead of edging out the out-of-domain training data as in fine-tuning.

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*8{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} \\ \hline
    FT-Full($d$) \hfill [BLEU] &37.7&59.2&54.5&34&46.8&90.8&53.8\\ \hline
    \hline
    \system{CL($d$)} &37.2&55.4&53.0&31.3&45.3&90.8&52.2\\ \hline
    \system{DDS($\vlambda_0, \vlambda_d$)} &36.7&56.2&54.3&33.5&46.8&91.8&53.2\\
    \system{MDAC($\vlambda_0, \vlambda_d$)} &36.7&58.2$^{**}$&55$^{*}$&33.3&46.6&91.9$^{***}$&53.6\\
     \hline
    \system{DDS($\vlambda_1, \vlambda_d$)} &36.4&56.3& 45.4& 32.6&44.5&43.1&43.1\\
    \system{MDAC($\vlambda_1, \vlambda_d$)} &37.2&55.1&51.8&32.8&44.9&85.4&51.2 \\ \hline
    \system{DDS($\vlambda_d, \vlambda_d$)} &37.0&59.6&53.7&30.2&45.7&91.6&53.0 \\
    \system{MDAC($\vlambda_d, \vlambda_d$)}&37.9&59.6&53.9&30.4&45.8&92.1&53.3 \\
    \hline
  \end{tabular}
  \caption{We report scores of each method with 6 target domains, and compare with full fine-tuning: each column corresponds to a distinct system. ($^{*}$) MDAC is significantly better than CL, fine-tuning and DDS with $p<0.05$. ($^{**}$) MDAC is significantly better than CL and DDS with $p<0.05$. ($^{***}$) MDAC is significantly better than CL, fine-tuning with $p<0.05$}
  \label{tab:da}
\end{table*}

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*8{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} \\ \hline \hline
    % Multidomain
    \system{Mixed-0}      & 35.3 & 54.1 & 52.5 & 31.9 & 44.9 & 89.5& 51.4 \\
    \system{Mixed-0.25} & 35.9 & 54.9 & 52.6 & 32.6 & 45 & 90.3& 51.9 \\
    \system{Mixed-0.5}   & 36.1 & 55.4 & 51.8 & 33.5 & 46.2 & 90 & 52.2 \\
    \system{Mixed-0.75} & 36.5 & 55 & 51.2 & 34 & 44.3 & 87.2& 51.7 \\
    \system{Mixed-1}     & 37.3  & 54.6 & 50.1 & 33.5 & 43.2 & 77.5& 49.4 \\
    \hline \hline
    \system{DDS($\vlambda_0, \vlambda_0$)} &37.3&55.1&51&33.5&43.4&90.8&51.8\\ 
    \system{MDAC($\vlambda_0, \vlambda_0$)}&35.5&55.5&53.7$^*$&32.3&45.6&91.1$^*$&52.3$^{**}$\\ 
    \hline \hline
    \system{DDS($\vlambda_1, \vlambda_0$)} &37.9&50.2&47.2&31.2&40.8&58.6&44.3\\
    \system{MDAC($\vlambda_1, \vlambda_0$)}&37.4&54.4&50.3&32.0&42.7&80.2&49.5\\
    \hline
  \end{tabular}
  \caption{We report the performance of our method MDAC and the work of \citet{Wang20balancing} DDS in multi-domain adaptation task. For a given line, all the columns \emph{correspond to the same multi-domain system}. ($^*$) MDAC is significantly better than \system{Mixed-$\alpha$} with $p<0.05$ and DDS. ($^{**}$) MDAC is significantly better than DDS with $p<0.05$.}
  \label{tab:multi-da}
\end{table*}

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*9{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} & \multicolumn{1}{|c|}{\domain{news}} \\ \hline \hline
    \system{MDAC($\vlambda_0, \vlambda_2$)}&36.2&-&-&-&46.1$^*$&-&-&- \\
    \system{DDS($\vlambda_0, \vlambda_2$)}&36.5&-&-&-&44.4&-&-&- \\
    \system{MDAC($\vlambda_0, \vlambda_2$)}&-&57.2$^*$&53.8$^*$&-&-&-&-&- \\
    \system{DDS($\vlambda_0, \vlambda_2$)}&-&56.1&52.6&-&-&-&-&- \\
    \system{MDAC($\vlambda_0, \vlambda_2$)}&-&-&-&32&-&90&-&- \\
    \system{DDS($\vlambda_0, \vlambda_2$)}&-&-&-&32.9&-&91.1&-&- \\ \hline
  \end{tabular}
  \caption{Non empty columns correspond to the target domains. For a given line, all the non empty columns \emph{correspond to the same bi-domain system}. ($^*$) MDAC is significantly better than DDS with $p<0.05$.}
  \label{tab:bi-da}
\end{table*}
According to Table \ref{tab:bi-da}, MDAC outperforms DDS in 2 over 3 settings including adapting to \domain{med-it} and \domain{law-bank} while being surpassed in \domain{talk-rel}, showing that MDAC is capable to adapt to a variable set of target domains.

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*9{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} & \multicolumn{1}{|c|}{\domain{news}} \\ \hline \hline
    \multicolumn{9}{|c|}{\sl Unsupervised domain adaptation} \\ \hline
    \system{Mixed-0}      & - & - & - & - & - & - & - & 22.4 \\
    \system{Mixed-0.25} & - & - & - & - & - & - & - & 22.7\\
    \system{Mixed-0.5}   & - & - & - & - & - & - & - & 23.6\\
    \system{Mixed-0.75} & - & - & - & - & - & - & - & 23.7\\
    \system{Mixed-1} & - & - & - & - & - & - & - & 23.5\\
    \hline \hline
    \system{MDAC($\vlambda_0, \vlambda_{news}$)} &-&-&-&-&-&-&-&23.1 \\
     \system{DDS($\vlambda_0, \vlambda_{news}$)} &-&-&-&-&-&-&-&23.3 \\ \hline\hline 
    \multicolumn{9}{|c|}{\sl 30 clusters} \\ \hline
    \system{MDAC($\vlambda^*, \vlambda_d$)}&36.3$^*$&56.9$^*$&51.3$^*$&33.2$^*$&43.4&89.4&51.8$^*$&-\\
     \system{DDS($\vlambda^*, \vlambda_d$)}&35.3&55.2&49.4&30.6&43.3&89.9&50.6&-\\ \hline
  \end{tabular}
  \caption{Non empty columns correspond to the target domains. For a given line, each non empty column correspond to one expert system. ($^*$)MDAC is significantly better than DDS.}
  \label{tab:unsupervised-da}
\end{table*}

@@

Most approaches to adaptive / dynamic data selection take inspiration from \citet{Bengio09curriculum}, where the notion of \emph{curriculum learning} (CL) is initially introduced. CL relies on the notion of ``easyness'' of a sample to schedule the presentation of the training data so that the easiest examples presented first and the hardest, last. CL has been shown to speed up learning. While the initial aim was to improve and speed up training, it has also proven useful for domain adaptive / multi-domain / multi-lingual MT, based on alternative definitions of ``easiness''.\fyTodo{Ajouter Grave}

@@

\section{Conclusion and future work}
\fyTodo{Write conclusion}

Contribution : generic framework for several settings. // comparison of several dynamic strategies.

Future work:
slow down learning during warm up or normalize reward; handle large number of classes;} 
too much importance to the dev ? randomize ?

>>>>

We now compare the best MDAC (initialized by $\lambda_0$) to fine-tuning. According to Table \ref{tab:da}, MDAC is outperformed by fine-tuning method in large domains including \domain{med} and \domain{law}. However, MDAC improves fine-tuning by approximately 1.2 BLEU in \domain{bank} and 1.0 BLEU in \domain{rel}. This indicates that for small domains, out-of-domain data can improve the generalization of the NMT model and MDAC is capable to exploit this advantage by balancing the in-domain and the out-of-domain training data instead of edging out the out-of-domain training data as in fine-tuning.

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*8{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} \\ \hline
    FT-Full($d$) \hfill [BLEU] &40.3&63.8&54.4&38.5&52.0&91.0&56.7\\ \hline
    \hline
    \system{CL($d$)} &40.2&60.2&53.7&36.5&51.1&91.1&55.5\\ \hline
    \system{DDS($\vlambda_0, \vlambda_d$)} &39.6&60.1&55.0&38.5&52.5&92.0&56.3\\
    \system{MDAC($\vlambda_0, \vlambda_d$)} &39.6&62.5$^{**}$&55.6$^{*}$&38.5&52.4&92$^{***}$&56.8\\
     \hline
    \system{DDS($\vlambda_1, \vlambda_d$)} &39.7&53.9&49.6&37.9&43.1&64.3&48.1\\
    \system{MDAC($\vlambda_1, \vlambda_d$)} &40.2&59.9&52.6&38.5&50.7&79.8&53.6\\ \hline
    \system{DDS($\vlambda_d, \vlambda_d$)} &39.9&63.9&54.5&35.4&51.2&91.8&56.1\\
    \system{MDAC($\vlambda_d, \vlambda_d$)}&40.6&63.9&54.5&35.6&51.3&92.3&56.4\\
    \hline
  \end{tabular}
  \caption{We report scores of each method with 6 target domains, and compare with full fine-tuning: each column corresponds to a distinct system. ($^{*}$) MDAC is significantly better than CL, fine-tuning and DDS with $p<0.05$. ($^{**}$) MDAC is significantly better than CL and DDS with $p<0.05$. ($^{***}$) MDAC is significantly better than CL, fine-tuning with $p<0.05$}
  \label{tab:da}
\end{table*}

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*8{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} \\ \hline \hline
    % Multidomain
    \system{Mixed-0} &38.6&59.3&53.7&37.3&51.0&90.4&55.1\\
    \system{Mixed-0.25}&38.9&59.6&53.3&37.6&50.5&90.6&55.1\\
    \system{Mixed-0.5}&39.0&60.2&52.5&38.5&51.9&90.3&55.4\\
    \system{Mixed-0.75}&39.4&59.9&51.9&38.8&50.0&87.6&54.6\\
    \system{Mixed-1}&40.3&59.5&49.8&36.4&49.0&80.0&52.5\\
    \hline \hline
    \system{DDS($\vlambda_0, \vlambda_0$)} &40.1&56.9&50.7&37.4&46.8&92.0&54.0\\ 
    \system{MDAC($\vlambda_0, \vlambda_0$)}&38.5&60.3$^{**}$&54.4$^*$&37.3&51.3$^{**}$&91.4$^*$&55.5$^{**}$\\ 
    \hline \hline
    \system{DDS($\vlambda_1, \vlambda_0$)} &40.6&55.5&48.0&36.2&46.9&60.1&47.9\\
    \system{MDAC($\vlambda_1, \vlambda_0$)}&40.2&59.3$^{**}$&51.0$^{**}$&36.9$^{**}$&48.6$^{**}$&80.7$^{**}$&52.8$^{**}$\\
    \hline
  \end{tabular}
  \caption{We report the performance of our method MDAC and the work of \citet{Wang20balancing} DDS in multi-domain adaptation task. For a given line, all the columns \emph{correspond to the same multi-domain system}. ($^*$) MDAC is significantly better than \system{Mixed-$\alpha$} with $p<0.05$. ($^{**}$) MDAC is significantly better than DDS with $p<0.05$.}
  \label{tab:multi-da}
\end{table*}

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*9{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} & \multicolumn{1}{|c|}{\domain{news}} \\ \hline \hline
    \system{DDS($\vlambda_0, \vlambda_2$)}&39.5&-&-&-&50.1&-&-&- \\
    \system{MDAC($\vlambda_0, \vlambda_2$)}&39.1&-&-&-&51.8$^*$&-&-&- \\
    \system{DDS($\vlambda_0, \vlambda_2$)}&-&60.8&53.3&-&-&-&-&- \\
    \system{MDAC($\vlambda_0, \vlambda_2$)}&-&61.9$^*$&54.5$^*$&-&-&-&-&- \\
    \system{DDS($\vlambda_0, \vlambda_2$)}&-&-&-&37.9&-&91.3&-&- \\ 
    \system{MDAC($\vlambda_0, \vlambda_2$)}&-&-&-&36.9&-&90.4&-&- \\
    \hline
  \end{tabular}
  \caption{Non empty columns correspond to the target domains. For a given line, all the non empty columns \emph{correspond to the same bi-domain system}. ($^*$) MDAC is significantly better than DDS with $p<0.05$.}
  \label{tab:bi-da}
\end{table*}
According to Table \ref{tab:bi-da}, MDAC outperforms DDS in 2 over 3 settings including adapting to \domain{med-it} and \domain{law-bank} while being surpassed in \domain{talk-rel}, showing that MDAC is capable to adapt to a variable set of target domains.

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*{10}{r|}} \cline{1-2} \cline{4-11}
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{news}}& \hfill &domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean}  \\ 
\cline{1-2} \cline{4-11}
    \multicolumn{2}{|c|}{\sl unseen domains} & &\multicolumn{8}{c|}{\sl 30 clusters} \\ 
\cline{1-2} \cline{4-11}
    \system{Mixed-0}      & 25.7 & &\system{DDS($\vlambda^*, \vlambda_d$)}&38.3&60.1&50.3&35.8&49.1&90.1&53.9\\
    \system{Mixed-0.25} & 25.8 & &\system{MDAC($\vlambda^*, \vlambda_d$)}&39.2$^*$&61.6$^*$&52.0$^*$&38.2$^*$&49.1&89.7&55.0$^*$\\ \cline{4-11}
    \system{Mixed-0.5}   &26.5\\
    \system{Mixed-0.75} &26.8\\
    \system{Mixed-1} &26.9 \\
    \cline{1-2}
     \system{DDS($\vlambda_0, \vlambda_{news}$)} &26.3 \\
     \system{MDAC($\vlambda_0, \vlambda_{news}$)} &26.3 \\
     \cline{1-2}
  \end{tabular}
  \caption{Non empty columns correspond to the target domains. For a given line, each non empty column correspond to one expert system. ($^*$)MDAC is significantly better than DDS.}
  \label{tab:unsupervised-da}
\end{table*}

@@

Most approaches to adaptive / dynamic data selection take inspiration from \citet{Bengio09curriculum}, where the notion of \emph{curriculum learning} (CL) is initially introduced. CL relies on the notion of ``easiness'' of a sample to schedule the presentation of the training data so that the easiest examples presented first and the hardest, last. CL has been shown to speed up learning. While the initial aim was to improve and speed up training, it has also proven useful for domain adaptive / multi-domain / multi-lingual MT, based on alternative definitions of ``easiness''.

@@
